Efficient Computation of Extended Surface Sources

### 2024-02-15 repro/papers/GEM22/submitR1/paper.tex
cleaned up

First, reverse-time propagation can be localized via ray theory,
within high-frequency asymptotic approximation. This step requires
some assumptions about the ray fields: all rays responsible for
significant energy in the receiver data must arrive from the source
surface, and must cross the source and receiver surfaces
transversally, that is, making non-zero angles. Also, the rays must
approach source and receiver surfaces from one side or the other
(``the inside''), so that locally the surfaces can be treated as
boundaries of propagation domains. Specifically, rays must emanate
from the source surface towards the inside (``incoming'' rays), and
arrive at the receiver surface from the inside (``outgoing'' rays).
With these assumptions, reverse-time propagation of the receiver data
closely approximates the acoustic fields near the source surface. Next
observe that for a causal field with a surface pressure source (with a
continuous pressure field across the surface), the source is
proportional to the jump in the (particle) velocity field. Moreover,
to leading order in frequency, the velocity field switches sign at the
surface, so the jump is just twice the sampled value on the
surface. Thus reverse-time propagation and reading off the velocity
field on the source surface yields a source that reproduces the
pressure data, within an asymptotically small error.
%Preconditioning amounts to a change of
%norm in the domain and range spaces of the modeling operator. In this
%case, the modfied norms are weighted $L^2$, and the weight operators
%map pressure to corresponding surface source on the source and
%receiver planes.

This is the essence of the approximate solution of the problem
\ref{eqn:esis} for penalty weight $\alpha=0$. To see how Krylov
iteration might be accelerated, note that time-reverse propagation of
the receiver pressure values may be represented by a time-reverse
source propagation, with the source constructed from the velocity
field at the receiver surface. Moreover, the reverse-time
source-to-pressure propagation is the transpose (adjoint) of the
forward-time source-to-pressure propagation - this is a very simple
version of the adjoint state construction \cite[]{Plessix:06}. So the
approximate inverse may be described as: conversion of pressure to source at the
receiver surface, followed by the adjoint of the modeling operator,
followed by the conversion of pressure to source at the source
surface. This sequence precisely describes the adjoint of the modeling
operator with respect to weighted norms in the spaces of source and
receiver data, with the weight operators being the
pressure-to-source operators. That this construction results in an
approximate inverse means that the modeling operator is approximately
unitary with respect to these weighted norms, which in turn implies
that a properly constructed preconditioned conjugate gradient algorithm should converge
rapidly.

FInally, the $\alpha=0$ case is not sufficient: the penalty operator
$A$ is an essential component of the nonlinear inverse propblem
\ref{eqn:esi}. As it turns out, $A$ commutes with the other operators
involved - approximately, but that is enough. Therefore the effect of
$A$ can be compensated with an easily-computed factor, whence the
preconditioned acceleration extended to the case $\alpha > 0$.

This pressure-to-source map is closely related to the ``hyperbolic
Dirichlet-to-Neumann'' operator that plays a prominent role in
photoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. \cite{HouSymes:EAGE16} demonstrated a
very similar preconditioner for Least Squares Migration, also for its
subsurface offset extension \cite[]{HouSymes:16}, motivated by
\cite{tenKroode:12}. These constructions also involve the
Dirichlet-to-Neumann operator. This concept also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.

%The obvious computation of the pressure-to-source map - prescribe the
%pressure, solve the wave equation with this boundary condition, read
%off the equivalent source - suffers from intrinsic numerical
%inaccuracy. I suggest an alternative computationally feasible
%approach, via economical short-distance wave propagation. Since the
%map is symmetric only in an approximate, asymptotic sense, it must be
%symmetrized for use as a Krylov preconditioner. I describe a
%symmetrization procedure that requires no further wave computations
%beyond those necessary to compute the action of the operator itself.

The discussion in this paper is formal and incomplete, in the sense
that some important mathematical underpinnings are only
sketched. I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
it mapped square integrable surface sources to square integrable sampled
data. This is not true in full generality: while the surface source
problem has distribution solutions, they are not generally square
integrable (finite acoustic field energy). Even if the solutions have
finite energy, they do not in general have well-defined
restrictions to lower-dimensional sets. In other words, the action of
the sampling
operator $P$ on the receiver surface is not well-defined for arbitrary
finite-energy acoustic fields. Thus the modeling operator envisioned above
is not be well-defined, strictly speaking.

This phenomenon is related to the ill-posedness of wave equations as
evolution equations in spatial variables, an observation attributed to
Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). A number of
authors have described precise forms of the ray conditions described
above, and shown how these conditions lead to the desired behaviour of
the modeling operator, that is, mapping of finite energy sources to
finite energy data, or equivalent properties
\cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}.
Elaboration of these mathematical details is beyond the scope of this
paper, which aims instead to explore the algorithmic consequences of
the mathematical structure implied by the nongrazing hypothesis.

@@

\begin{figure}
  \centering
  \subfloat[\label{fig:drecplh0}]{\includegraphics[width=0.32\textwidth]{drecplh0.pdf}}
  \subfloat[\label{fig:dsrcphh0}]{\includegraphics[width=0.32\textwidth]{dsrcphh0.pdf}}
  \subfloat[\label{fig:dfwdplh0}]{\includegraphics[width=0.32\textwidth]{dfwdplh0.pdf}}
  \caption{Data generated using configuration in Figure
    \ref{fig:olensrays0}. (a) traces from point source at $(3500,3500)$
    with $[1, 2, 7.5, 12.5]$ Hz zero phase trapezoidal bandpass
    filter wavelet, delayed by $0.5$ s. (b) Equivalent extended source on
    the source surface at depth $z_s=3000$ m. (c) Gather generated by
    equivalent source shown in (b). Color scale is same for (a) and (c).}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{ooplrays0.pdf}
  \caption{Bulk modulus generating diving waves. Color scale unit is
    GPa. Orange horizontal line is source surface at depth $z_s= 500$
    m, yellow horizontal
    line is receiver surface at depth $z_r=100$ m. Point source location for data
    generation at $(500, 10000)$ m indicated with star. Overlain with rays from point
    source to receiver surface.}
  \label{fig:ooplrays0}
\end{figure}

@@

Time reversal as found in the literature on photoacoustic tomography
must be localized to apply to the surface source inversion problem,
since there is no measurement surface surrounding the region of
propagation. It must also be modified to produce a source, rather than
a restriction of the pressure field. The assumption of incoming rays at the source surface and
outgoing rays at the receiver surface makes both modifications  
possible.

The construction takes place in three stages: first, an
acoustic field with high-frequency energy propagating along an
incoming ray family is identified as the causal solution of an inhomogeneous boundary value problem {\em locally and on
  the inside}. The boundary value is simply the restriction of the
acoustic field to the surface. Second, the same causal solution is
identified as the causal solution of an acoustic problem with source,
singular on the surface in the fashion of equation
\ref{eqn:awepm}. The source amplitude is proportional to 
the velocity field sampled on the surface. Third, this identification is used twice, first to
propagate the receiver data backwards in time as the solution of an
anti-causal problem,  the velocity field of which is sampled on the
source surface; then the causal problem is solved with the
source built from the velocity samples. This solution must be
identical to the field generating the data, along the ray fields
carrying all of the high-frequency energy. Therefore the reconstructed
source is an approximate solution of the problem \ref{eqn:esis} for
$\alpha=0$.

@@

In this subsection $(p,\bv)$ is a solution of the homogeneous acoustic system
\begin{eqnarray}
\label{eqn:awebase}
  \frac{1}{\kappa}\frac{\partial p}{\partial t} & = & - \nabla \cdot \bv, \nonumber \\
  \rho\frac{\partial \bv}{\partial t} & = & - \nabla p,
\end{eqnarray}
The source surface $\Sigma$ is given a choice of unit normal field ${\bf
  n}$. Rays carrying
high-frequency energy in $(p,\bv)$ pass through $\Sigma$, and
intersect $\Sigma$ only at times
$>0$. Also, the velocity
vectors of all such rays have a negative ${\bf n}$ component where
they cross $\Sigma$. Construct a time interval $[0,\tau]$ and
a region in space $\Omega^+$ containing $\Sigma$ so that $\Sigma$ forms
part of the boundary $\partial \Omega^+$ of $\Omega^+$, ${\bf n}$ is the
outward unit normal to $\Omega^+$ on $\Sigma$, and the energetic
rays cross $\partial \Omega^+$ {\em only} at
points of $\Sigma$ in the time interval $[0,\tau]$ - that is, in time
less than $\tau$ the rays do not reach other parts of $\partial
\Omega^+$. Note that the rays are {\em incoming} to $\Omega^+$ in the terminology
introduced earlier. See Figure \ref{fig:omegarays}.

@@

Consequent to this ray geometry, conclude that the field $(p,\bv)$  is asymptotically negligible both
throughout $\Omega^+$ at $t=0$ and at all points of 
$\partial \Omega^+$ except for $\Sigma$, for $0\le t \le \tau$. So $(p,\bv)$ differs
negligibly from the solution of the boundary value problem 
\begin{eqnarray}
\label{eqn:aweloc}
  \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} & = & - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla
                                                    \tilde{p}, \\
  \tilde{p}(\bx,t) & = & p(\bx,t) \mbox{ for } \bx \in \Sigma, 0 \le t \le 
  \tau,\\
  \tilde{p}(\bx,t) & = & 0 \mbox{ for } \bx \in \partial \Omega^+ \setminus \Sigma, 0 \le t \le
 \tau.
\end{eqnarray}

@@

The fields involved in the surface source extension are causal and
source-generated, rather than solutions of boundary value problems
such as the system \ref{eqn:aweloc}. To transition to a source
representation, begin by extending the domain $\Omega^+$, lying on the
incoming side of the surface $\Sigma$, to a full neighborhood $\Omega$
of sigma
by adding a domain $\Omega^-$ on the outside (see Figure
\ref{fig:omegarays}). Once again, construct $\Omega^-$ so that none of
the rays carrying high-frequency energy cross the boundary of
$\Omega^-$ except at $\Sigma$. Extend the field $(\tilde{p},\tilde{\bv})$ to the
full neigborhood $\Omega$ by solving the system \ref{eqn:aweloc} with
$\Omega^+$ replaced by $\Omega^-$. Thus the extended
field is pieced together out of causal fields solving the boundary value
problem on either side of the boudary, with {\em the same} boundary
condition, that is, the trace of $p$ on $\Sigma$. Thus the extended
field is continuous in $p$. it is not however continuous in
$\bv$, as will be seen shortly. More precisely, the normal component
$\bv\cdot {\bf n}$
has a discontinuity, while the tangential components are continuous.
The jump in
$\bv\cdot {\bf n}$ contributes a delta function to the gradient: as a
field in the whole neigborhood $\Omega$, $(\tilde{p},\tilde{\bv})$
solves
\begin{eqnarray}
\label{eqn:awebase}
  \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} & = & [\bv \cdot {\bf n}]_{\Sigma}\delta_{\Sigma}
                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p},\\
  \tilde{p}(\bx,t) & = & 0 \mbox{ for } \bx \in \partial \Omega, 0 \le t \le 
  \tau,\\
  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega^+, t , \\
  \tilde{\bv} & = & {\bf 0} \mbox{ for } \bx \in \Omega^+, t = 0.       
\end{eqnarray}
in which $\delta_{\Sigma}$ is the delta function on $\Sigma$. The
square bracket denotes the jump in the quantity enclosed across
$\Sigma$, from the outside to the inside. Since ${\bf n}$ points from
the inside to the outside, this convention contributes another minus
sign.

@@

There is a downside, namely the apparent necessity of constructing the
solution in $\Omega^-$ so that the jump of $\bv \cdot {\bf n}$ may be
computed. That is actually not necessary, however.
This is best seen by analyzing a generic geometric optics
component associated to the phase function $\psi^+$ in $\Omega^+$ and its family of
rays. The acoustic field is assembled by a sum (in the continuum, an
integral) over such solutions. A particular summand looks like
\begin{equation}
  \label{eqn:go}
   a(\bx)e^{i\omega (t - \psi(\bx))}.
\end{equation}
for $\bx \in \Omega^+$.
The phase $\psi$ solves the eikonal equation, the amplitude $a$ the
related transport equation. The restriction of $p$
to $\Sigma$ determines $\psi$ (and $a$) restricted to $\Sigma$, essentially by
Fourier synthesis. Since the eikonal
equation is even in $\nabla \psi$,  and $\psi$ restricted to $\Sigma$
determines the tangential derivatives of $\psi$, the normal component
of $\nabla \psi$ is determined up to sign by the boundary values of
$p$. The spatial component of a ray $X(t)$ associated to $\psi$ solves
the ordinary differential equation
\begin{equation}
  \label{eqn:ray}
  \frac{dX}{dt} = \nabla \psi(X).
\end{equation}
The normal component of this ray must be negative (in
$\Omega^+$) - this is the incoming condition, and guarantees that the
ray does not intersect $t=0$ in $\Omega^+$ near $\Sigma$, justifying the initial
condition in \ref{eqn:aweloc}. Similarly, the normal component must be
positive in $\Omega^-$ near $\Sigma$. But both must have the same magnitude, since
$\psi$ solves the eikonal equation - so they are negatives of each
other. For causal solutions such as $(\tilde{p},\tilde{\bv})$, the
Newton's law equation in \ref{eqn:aweloc} is equivalent to
\begin{equation}\label{eqn:newt}
  \tilde{\bv}(\bx,t) = -\frac{1}{\rho}\int_{-\infty}^t ds \nabla \tilde{p}(\bx,s),
\end{equation}
except at $\Sigma$.
it follows that the normal components of $\tilde{\bv}$ have limits at
point of $\Sigma$ 
that are opposite in sign: for $\bx \in \Sigma$,
\[
  \lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf
    n}(\bx+\zeta {\bf n})) = 
-\lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf
  n}(\bx -\zeta {\bf n}))
\]
whence
\begin{equation}
[\tilde{\bv} \cdot {\bf n}] = 2  \label{eqn:vjump} \lim_{\zeta
  \rightarrow 0} (\tilde{\bv}(\bx,t) \cdot {\bf n}(\bx -\zeta {\bf
  n}))
\approx 2 \bv \cdot {\bf n}
\end{equation}
That is, at least over the short time interval $0 \le t \le \tau$,
$(p,\bv)$ is asymptotically approximated in $\Omega^+$ by
$(\tilde{p},\tilde{\bv})$, which solves
\begin{eqnarray}
\label{eqn:awesrc}
  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & f \delta_{\Sigma}
                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p},\\
  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega, t=0, \\
  \tilde{\bv} & = & {\bf 0} \mbox{ for } \bx \in \Omega, t = 0.       
\end{eqnarray}
with $f = 2 \bv \cdot {\bf n}$ on $\Sigma$.

This result appears limited by the short time assumption (on $\tau$)
and the corresponding spatial localization to $\Omega^+$. However
neither of these apparent limitations is real. The short time
assumption may be relaxed by windowing the fields to short time
windows then adding up the results. The approximating fields
$(\tilde{p}, \tilde{\bv})$ are themselves approximated by ray
theory solutions. The assumptions on rays crossing source and receiver
surfaces are global, and extend the approximation \ref{eqn:awesrc}
globally. Thus the last two conditions may be replaced by the basic
causality condition
\begin{eqnarray}
  \label{eqn:awesrcmod}
  ... & = & ...\\
  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \\
  \tilde{\bv} & = & {\bf 0} \mbox{ for } t \ll 0.       
\end{eqnarray}
The effect of this change on the receiver surface traces of $\tilde{p}$ and
$\tilde{\bv}$ is negligible.

>>>>

First, reverse-time propagation can be localized via ray theory,
within high-frequency asymptotic approximation. This step requires
some assumptions about the ray fields: all rays responsible for
significant energy in the receiver data must arrive from the source
surface, and must cross the source and receiver surfaces
transversally, that is, making non-zero angles. The surfaces must be
oriented, with a choice of (``outward'') unit normal vector. The
energetic rays must emanate
from the source surface inwards (``incoming'' rays), and
arrive at the receiver surface outwards (``outgoing'' rays). Then both surfaces
can be treated locally as
boundaries of propagation domains. 
With these assumptions, reverse-time propagation of the receiver data
closely approximates the acoustic fields near the source
surface. Next, I show that for a causal field with a surface pressure source (with a
continuous pressure field across the surface), the source amplitude is
proportional to the jump in the (particle) velocity field. Moreover,
to leading order in frequency, the velocity field switches sign at the
surface, so the jump is just twice the sampled value on the
surface. Thus constructing a source from the veloity trace on the
receiver surface, reverse-time propagation of this source, and reading off the velocity
field on the source surface yields a source that reproduces the
pressure data, within an asymptotically small error.

This is the essence of the approximate solution of the problem
\ref{eqn:esis} for penalty weight $\alpha=0$. To see how Krylov
iteration might be accelerated, I note that reverse-time propagation of
the receiver surface source to pressure on the source surface is the transpose (adjoint) of the
forward-time source-to-pressure propagation - this is a very simple
version of the adjoint state construction \cite[]{Plessix:06}. So the
approximate inverse may be described as: conversion of pressure to source at the
receiver surface, followed by application of the modeling operator adjoint,
followed by the conversion of pressure to source at the source
surface. This sequence precisely describes the adjoint of the modeling
operator with respect to weighted norms in the spaces of source and
receiver data, with the weight operators being the
pressure-to-source operators on the two surfaces. That this construction results in an
approximate inverse means that the modeling operator is approximately
unitary with respect to these weighted norms, which in turn implies
that a properly constructed preconditioned conjugate gradient algorithm should converge
rapidly.

Finally, the $\alpha=0$ case is not sufficient: the penalty operator
$A$ is an essential component of the nonlinear inverse propblem
\ref{eqn:esi}. As it turns out, $A$ commutes with the other operators
involved - approximately, but that is enough. Therefore the effect of
$A$ can be compensated with an easily-computed factor, whence the
preconditioned acceleration extended to the case $\alpha > 0$.

This pressure-to-source map is closely related to the ``hyperbolic
Dirichlet-to-Neumann'' operator that plays a prominent role in
photoacoustic tomography and other wave inverse problems
\cite[]{Rachele:00,StefUhl:05}. In this terminology, ``Dirichlet''
refers to pressure restricted to a surface, and ``Neumann'' to the
normal derivative of pressure on that surface. Since the latter is
proportional to normal particle acceleration thanks the equations of motion,
and normal particle velocity is proportion to source, the relation to the
pressure-to-source map as described in this paper is clear. A number
of authors have described preconditioners for Least Squares Migration
with subsurface offset extension in which the Dirichlet-to-Neumann or
pressure-to-source map plays a central role
\cite[]{tenKroode:12,HouSymes:15,Herve2017}. It also turns up in hidden
form in the work of Yu Zhang and collaborators on true amplitude
migration
\cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09}.

The discussion in this paper is formal and incomplete, in the sense
that some important mathematical underpinnings are only
sketched. The transport of high-frequency energy along rays is a
crucial ingredient in the constructions explained here. Of course,
such transport only occurs in material parameter models that vary
slowiy in space on the wavelength scale, and even then only
approximately. Rather than provide explicit
smoothness constraints and asymptotic error estimates, I will use
these concepts in an intuitive way, using ``approximate'' and
``asymptotic'' and the symbol ``$\approx$'' to suggest asymptotically
negligible errors not explicitly assessed, and refer to such errors as
``negligible''.

I will treat the modeling operator $PL[{\bf c}]^{-1}$ as if
it mapped square integrable surface sources to square integrable sampled
data. This is not true in full generality: while the surface source
problem has distribution solutions, they are not generally square
integrable (finite acoustic field energy). Even if the solutions have
finite energy, they do not in general have well-defined
restrictions to lower-dimensional sets. In other words, the action of
the sampling
operator $P$ on the receiver surface is not well-defined for arbitrary
finite-energy acoustic fields. Thus the modeling operator envisioned above
is not be well-defined, strictly speaking.
This phenomenon is related to the ill-posedness of wave equations as
evolution equations in spatial variables, an observation attributed to
Hadamard (see \cite{CourHil:62}, Chapter 6, section 17). A number of
authors have described precise forms of the ray conditions described
above, and shown how these conditions lead to the desired behaviour of
the modeling operator, that is, mapping of finite energy sources to
finite energy data, or equivalent properties
\cite{Payn:75,Symes:83,Lasi:86,LasLionsTrig:86,Lasi:87, BaoSy:91b}.
Elaboration of these mathematical details is beyond the scope of this
paper, which aims instead to explore the algorithmic consequences of
the mathematical structure implied by the nongrazing hypothesis.

@@

\begin{figure}
  \centering
  \subfloat[\label{fig:drecplh0}]{\includegraphics[width=0.32\textwidth]{drecplh0.pdf}}
  \subfloat[\label{fig:dhshh0}]{\includegraphics[width=0.32\textwidth]{dhshh0.pdf}}
  \subfloat[\label{fig:dfwdplh0}]{\includegraphics[width=0.32\textwidth]{dfwdplh0.pdf}}
  \caption{Data generated using configuration in Figure
    \ref{fig:olensrays0}. (a) traces from point source at $(3500,3500)$
    with $[1, 2, 7.5, 12.5]$ Hz zero phase trapezoidal bandpass
    filter wavelet, delayed by $0.5$ s. (b) Equivalent extended source on
    the source surface at depth $z_s=3000$ m. (c) Gather generated by
    equivalent source shown in (b). Color scale is same for (a) and (c).}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{ooplrays0.pdf}
  \caption{Bulk modulus generating diving waves. Color scale unit is
    GPa. Orange horizontal line is source surface at depth $z_s= 500$
    m, yellow horizontal
    line is receiver surface at depth $z_r=100$ m. Point source location for data
    generation at $(500, 10000)$ m indicated with star. Overlain with rays from point
    source to receiver surface. Outward normal points up at both source
    and receiver surfaces, so rays are incoming at source
    surface, outgoing at receiver surface.}
  \label{fig:ooplrays0}
\end{figure}

@@

The time reversal construction, as found in the literature on photoacoustic tomography,
must be localized to apply to the surface source inversion problem,
since there is no measurement surface surrounding the region of
propagation. It must also be modified to produce a source, rather than
a slice of the pressure field. The assumption of incoming rays at the source surface and
outgoing rays at the receiver surface makes both modifications  
possible.

The construction takes place in three stages: first, an acoustic field
with high-frequency energy propagating along an incoming ray family is
identified as the causal solution of an inhomogeneous boundary value
problem {\em locally and on the inside}. The boundary value is simply
the restriction of the pressure field to the surface. Second, the same
causal solution is identified as the causal solution of an acoustic
problem with source, singular on the surface in the fashion of
equation \ref{eqn:awepm}. The source amplitude is proportional to the
velocity field sampled on the surface. Third, this identification is
used twice, first to propagate the receiver data backwards in time as
the solution of an anti-causal problem with source on the receiver
surface; the velocity field of which is sampled on the source surface;
then the causal problem is solved with the source built from the
velocity samples. This solution must asymptotically approximate the
field generating the data, along the ray fields carrying all of the
high-frequency energy. Therefore the reconstructed source is an
approximate solution of the problem \ref{eqn:esis} for $\alpha=0$.

@@

In this subsection $(p,\bv)$ is a solution of the homogeneous acoustic system
\begin{eqnarray}
\label{eqn:awefree}
  \frac{1}{\kappa}\frac{\partial p}{\partial t} & = & - \nabla \cdot \bv, \nonumber \\
  \rho\frac{\partial \bv}{\partial t} & = & - \nabla p,
\end{eqnarray}
The source surface $\Sigma$ is given a choice of unit normal field ${\bf
  n}$. Rays carrying
high-frequency energy in $(p,\bv)$ pass through $\Sigma$, and
intersect $\Sigma$ only at times
$>0$. Also, the velocity
vectors of all such rays have a negative ${\bf n}$ component where
they cross $\Sigma$. Construct a time interval $[0,\tau]$ and
a region in space $\Omega^+$ containing $\Sigma$ so that $\Sigma$ forms
part of the boundary $\partial \Omega^+$ of $\Omega^+$, ${\bf n}$ is the
outward unit normal to $\Omega^+$ on $\Sigma$, and the energetic
rays cross $\partial \Omega^+$ {\em only} at
points of $\Sigma$ in the time interval $[0,\tau]$ - that is, in time
less than $\tau$ the rays do not reach other parts of $\partial
\Omega^+$. Note that the rays are {\em incoming} to $\Omega^+$ in the terminology
introduced earlier. See Figure \ref{fig:omegarays}.

@@

Consequent to this ray geometry, conclude that the field $(p,\bv)$  is asymptotically negligible both
throughout $\Omega^+$ at $t=0$ and at all points of 
$\partial \Omega^+$ except for $\Sigma$, for $0\le t \le \tau$. So $(p,\bv)$ differs
negligibly from the solution of the boundary value problem 
\begin{eqnarray}
\label{eqn:aweloc}
  \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} & = & - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla
                                                    \tilde{p}, \nonumber \\
  \tilde{p}(\bx,t) & = & p(\bx,t) \mbox{ for } \bx \in \Sigma, 0 \le t \le 
  \tau, \nonumber \\
  \tilde{p}(\bx,t) & = & 0 \mbox{ for } \bx \in \partial \Omega^+ \setminus \Sigma, 0 \le t \le
 \tau.
\end{eqnarray}

@@

The fields involved in the surface source extension are causal and
source-generated, rather than solutions of boundary value problems
such as the system \ref{eqn:aweloc}. To transition to a source
representation, begin by extending the domain $\Omega^+$, lying on the
incoming side of the surface $\Sigma$, to a full neighborhood $\Omega$
of sigma
by adding a domain $\Omega^-$ on the outside (see Figure
\ref{fig:omegarays}). Once again, construct $\Omega^-$ so that none of
the rays carrying high-frequency energy cross the boundary of
$\Omega^-$ except at $\Sigma$. Extend the field $(\tilde{p},\tilde{\bv})$ to the
full neigborhood $\Omega$ by solving the system \ref{eqn:aweloc} with
$\Omega^+$ replaced by $\Omega^-$. Thus the extended
field is pieced together out of causal fields solving the boundary value
problem on either side of the boudary, with {\em the same} boundary
condition, that is, the trace of $p$ on $\Sigma$. Thus the extended
field is continuous in $p$. it is not however continuous in
$\bv$, as will be seen shortly. More precisely, the normal component
$\bv\cdot {\bf n}$
has a discontinuity, while the tangential components are continuous.
The jump in
$\bv\cdot {\bf n}$ contributes a delta function to the gradient: as a
field in the whole neigborhood $\Omega$, $(\tilde{p},\tilde{\bv})$
solves
\begin{eqnarray}
\label{eqn:awebase}
  \frac{1}{\kappa}\frac{\partial \tilde{p}}{\partial t} & = & [\bv \cdot {\bf n}]_{\Sigma}\delta_{\Sigma}
                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p}, \nonumber \\
  \tilde{p}(\bx,t) & = & 0 \mbox{ for } \bx \in \partial \Omega, 0 \le t \le 
  \tau, \nonumber \\
  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega^+, t , \nonumber \\
  \tilde{\bv} & = & {\bf 0} \mbox{ for } \bx \in \Omega^+, t = 0.       
\end{eqnarray}
The square bracket denotes the jump in the quantity enclosed across
$\Sigma$, from the outside to the inside. Since ${\bf n}$ points from
the inside to the outside, this convention contributes another minus
sign.

@@

There is a downside, namely the apparent necessity of constructing the
solution in $\Omega^-$ so that the jump of $\bv \cdot {\bf n}$ may be
computed. That is actually not necessary, however.
This is best seen by analyzing a generic geometric optics
component associated to the phase function $\psi$ in $\Omega^+$ and its family of
rays. The acoustic field is assembled by a sum (in the continuum, an
integral) over such solutions. A particular summand looks like
\begin{equation}
  \label{eqn:go}
   a(\bx,t)e^{i\omega (t - \psi(\bx))}.
\end{equation}
for $\bx \in \Omega^+$.
The phase $\psi$ solves the eikonal equation, the amplitude $a$ the
related transport equation. The restriction of $p$
to $\Sigma$ determines $\psi$ (and $a$) restricted to $\Sigma$, essentially by
Fourier synthesis. Since the eikonal
equation is even in $\nabla \psi$,  and $\psi$ restricted to $\Sigma$
determines the tangential derivatives of $\psi$, the normal component
of $\nabla \psi$ is determined up to sign by the boundary values of
$p$. The spatial component of a ray $X(t)$ associated to $\psi$ solves
the ordinary differential equation
\begin{equation}
  \label{eqn:ray}
  \frac{dX}{dt} = \nabla \psi(X).
\end{equation}
The normal component of this ray must be negative (in
$\Omega^+$) - this is the incoming condition, and guarantees that the
ray does not intersect $t=0$ in $\Omega^+$ near $\Sigma$, justifying the initial
condition in \ref{eqn:aweloc}. Similarly, the normal component must be
positive in $\Omega^-$ near $\Sigma$. But both must have the same magnitude, since
$\psi$ solves the eikonal equation - so they are negatives of each
other. For causal solutions such as $(\tilde{p},\tilde{\bv})$, the
Newton's law equation in \ref{eqn:aweloc} is equivalent to
\begin{equation}\label{eqn:newt}
  \tilde{\bv}(\bx,t) = -\frac{1}{\rho}\int_{-\infty}^t ds \nabla \tilde{p}(\bx,s),
\end{equation}
except at $\Sigma$.
it follows that the normal components of $\tilde{\bv}$ have limits at
point of $\Sigma$ 
that are opposite in sign: for $\bx \in \Sigma$,
\[
  \lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx+\zeta {\bf n}(\bx),t) \cdot {\bf
    n}(\bx)) = 
-\lim_{\zeta \rightarrow 0} (\tilde{\bv}(\bx-\zeta {\bf n}(\bx),t) \cdot {\bf
  n}(\bx))
\]
whence
\begin{equation}
   \label{eqn:vjump}
   [\tilde{\bv} \cdot {\bf n}] = 2 \lim_{\zeta
  \rightarrow 0} (\tilde{\bv}(\bx -\zeta {\bf
  n}(\bx),t) \cdot {\bf n}(\bx))
\approx 2 \bv(\bx) \cdot {\bf n}(\bx)
\end{equation}
That is, at least over the short time interval $0 \le t \le \tau$,
$(p,\bv)$ is asymptotically approximated in $\Omega^+$ by
$(\tilde{p},\tilde{\bv})$, which solves
\begin{eqnarray}
\label{eqn:awesrc}
  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & h \delta_{\Sigma}
                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p}, \nonumber \\
  \tilde{p} & = & 0 \mbox{ for } \bx \in \Omega, t=0, \nonumber \\
  \tilde{\bv} & = & {\bf 0} \mbox{ for } \bx \in \Omega, t = 0.       
\end{eqnarray}
with $h = 2 \bv \cdot {\bf n}$ on $\Sigma$.

This result appears limited by the short time assumption (on $\tau$)
and the corresponding spatial localization to $\Omega^+$. However
neither of these apparent limitations is real. The short time
assumption may be relaxed by windowing the fields to short time
windows then adding up the results. The approximating fields
$(\tilde{p}, \tilde{\bv})$ are themselves approximated by ray
theory solutions. The assumptions on rays crossing source and receiver
surfaces are global, and extend the approximation \ref{eqn:awesrc}
globally. Thus the last two conditions may be replaced by the basic
causality condition
\begin{eqnarray}
  \label{eqn:awesrcmod}
  \frac{1}{\kappa}\frac{\partial\tilde{p}}{\partial t} & = & h \delta_{\Sigma}
                                                      - \nabla \cdot \tilde{\bv}, \nonumber \\
  \rho\frac{\partial \tilde{\bv}}{\partial t} & = & - \nabla \tilde{p}, \nonumber \\
  \tilde{p} & = & 0 \mbox{ for }  t \ll 0, \nonumber \\
  \tilde{\bv} & = & {\bf 0} \mbox{ for } t \ll 0.       
\end{eqnarray}
The effect of this change on the receiver surface traces of $\tilde{p}$ and
$\tilde{\bv}$ is negligible.

An example has already appeared in Figure \ref{fig:dhshh0}. In this
example, the field $(p, \bv)$ is the point source solution with the
point source depicted in Figure \ref{fig:olensrays0}, located at
$(3500, 3500)$ m. The role of $\Sigma$ is played by the source surface $\Sigma_s=
\{z=z_s=3000 \mbox{ m}\}$, and near this surface the field solves the source-free
system \ref{eqn:awefree}. The ``inside'' in this example is the region
$z > z_s$, and the outward unit normal is therefore ${\bf n}=(0,0,-1)^T$.
Thus $(p,\bv)$ should be asymptotically close to the solution
$(\tilde{p},\tilde{\bv})$ of the
system \ref{eqn:awesrc} in the region $z< z_s$, at least at locations
connected by the ray field shown in Figure \ref{fig:olensrays0} to the
source surface, provided that the source amplitude $h$ is chosen $=
-2v_z$ on $\Sigma_s=\{z=z_s\}$. Figure \ref{fig:dhshh0} displays exactly this $h$, and
\ref{fig:dfwdplh0} the traces of $\tilde{p}$ on the
  receiver surface $z=z_r=1000$ m, which satisfies the ray condition. The extremely
close match between Figures \ref{fig:drecplh0} (traces of $p$ on $\Sigma_r$) and
\ref{fig:dfwdplh0} is consistent with the conclusion reached here.
